```{r, echo=FALSE}
library(caret, quietly = TRUE)
library(rpart, quietly = TRUE)
library(randomForest, quietly = TRUE)
library(rattle, quietly = TRUE)
```
---
title: "Practical Machine Learning - Prediction Assignment Writeup"
output:
  html_document:
    highlight: tango
    theme: journal
self_contained: no
keep_md: yes
---
```{r, echo=FALSE, cache = TRUE}
#Function for converting data types of multiple columns [6]
convert.magic <- function(obj, type, cols) {
    FUN1 <- switch(type, character = as.character, numeric = as.numeric, factor = as.factor)
    obj[,cols] <- lapply(obj[,cols], FUN1) 
    as.data.frame(obj)
}
#Download data. This may take a little while (about 35 seconds on my laptop.)
if (!file.exists("pml-training.csv")) {
    download.file(url="https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv",
                  destfile="pml-training.csv")
}
if (!file.exists("pml-testing.csv")) {
    download.file(url="https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv",
                  destfile="pml-testing.csv")
}
```
### Building a machine learning algorithm to predict activity quality from activity monitors  

#### Background  
The data for this project comes from this source: http://groupware.les.inf.puc-rio.br/har. This data contains telemetry from accelerometers on the belt, forearm, arm, and dumbell of 6 participants on a Human Activity Recognition project. The participants were asked to perform barbell lifts correctly and incorrectly in 5 different ways. These 5 ways are quantified under the `classe` variable of the training data set: exactly according to the specification (classe A), throwing the elbows to the front (classe B), lifting the dumbbell only halfway (classe C), lowering the dumbbell only halfway (classe D) and throwing the hips to the front (classe E).   
The goal of our course project is to build a machine learning algorithm to predict the manner in which a participant did the exercise.  


### I. How the prediction model was built  

#### I.1. Data Exploration  
While loading the training and testing data, we converted all blanks, spaces, string NAs and #DIV/0! to true NA values:  
```{r cache=TRUE}
trainingRaw <- read.csv("pml-training.csv", header=TRUE, na.strings = c("", " ", "NA", "#DIV/0!"))
testingRaw <- read.csv("pml-testing.csv", header=TRUE, na.strings = c("", " ","NA", "#DIV/0!"))
```
By looking at `summary(trainingRaw)` (output omitted) we realized there were many variables which had many NA values. There were a total of `r length(which(colSums(is.na(trainingRaw))>500))` variables that had more than 500 rows with NAs. Out of the 60 variables left, the first 7 were metadata only (e.g. *user_name*, *raw_timestamp_part_1*, *cvtd_timestamp*, etc.) We also realized that the rows with `new_window == "yes"` seem to contain summary data but only for the columns that normally have NA values (e.g. *kurtosis_*, *skweness_*, *max_*, *min_*, etc.)

#### I.2. Data Cleaning and Preparation  
Based on the knowledge gained during data exploration, we proceeded to clean and prepare the data in the following manner:  
    1. Removed all variables with more than 500 rows with NA values.  
    2. Filtered rows to only those that have new_window == "no".  
    3. Removed all 7 metadata columns.  
    4. Changed all variable data types to numeric (with the exception of classe variable.)  
    5. Split (0.7) the clearn training set into training and testing subsets based on the classe variable.  
```{r cache=TRUE}
# Remove columns with more than 500 NA values
trainingClean <- trainingRaw[!colSums(is.na(trainingRaw)) > 500]
# Filter rows to only those that have new_window == "no"
trainingClean <- subset(trainingClean, subset = (new_window == "no"))
# Remove metadata columns
trainingClean <- trainingClean[,-(1:7),drop=FALSE]
# Change all data types to numeric (with the exception of classe)
trainingClean <- convert.magic(trainingClean,type = "numeric",c(1:52))
# Split clean raw training set into training and testing sets based on classe
trainIndex = createDataPartition(trainingClean$classe, p = 0.70, list=FALSE)
trainingClean.training = trainingClean[trainIndex,]
trainingClean.testing = trainingClean[-trainIndex,]
```  

#### I.3. Algorithm Selection  
We proceeded with our analysis by trying two different machine learning algorithms:  

1. **Decision tree using 10-fold cross validation**:  
```{r cache=TRUE}
set.seed(123654) # Set a seed for reproducibility
train_control <- trainControl(method = "cv", number = 10)
modFit.rpart <- train(classe ~., method = "rpart", data = trainingClean.training, trControl = train_control)
# Show decision tree on a fancy format
fancyRpartPlot(modFit.rpart$finalModel)  
```  

2. **Random forest**:  
Please note that in random forests, there is no need for cross-validation or a separate test set to get an unbiased estimate of the test set error. It is estimated internally, during the run. [2]
```{r cache=TRUE}
modFit.rf <- randomForest(classe ~ ., data = trainingClean.training)
```
We also tried using a **boosting with trees** algorithm but did not include it on this report as the runtime was very long and it *did not improve the accuracy above what random forest already provided*. Code provided below for reference only.
```{r eval=FALSE}
modFit.gbm <- train(classe ~ ., method = "gbm", data = trainingClean.training, verbose = FALSE)
modFit.gbm
```  


### II. Cross validation  
In the previous section we performed 10-fold cross validation for the decision tree algorithm and the test set error was estimated internally during the run of Random forest. We'll now proceed to compare accuracy between these two algorithms.  

#### II.1. Algorithm Comparison  

1. **Decision tree using 10-fold cross validation**:  
We predict `classe` for the clean testing set using our decision tree model. By taking a quick look at the results we foresee the accuracy must be low as the model doesn't classify any observation under `classe == "D"`. As we can see the actual accuracy is `r round(max(modFit.rpart$results[2]),4)` which is basically the same as a coin flip.
```{r cache=TRUE}
pred.rpart <- as.vector(predict(modFit.rpart, trainingClean.testing))
unique(pred.rpart)
max(modFit.rpart$results[2])
```  

2. **Random forest**:  
We predict `classe` for the clean testing set using our random forest model. By looking at the confusion matrix class error estimates we can see that this model is better than the decision tree model.
```{r cache=TRUE}
pred.rf <- as.vector(predict(modFit.rf, trainingClean.testing))
modFit.rf$confusion
confusionMatrix(pred.rf, trainingClean.testing$classe)$overall[1]
```  
The accuracy is `r confusionMatrix(pred.rf, trainingClean.testing$classe)$overall[1]`. We will use this model onwards to calculate the out of sample error and predict the `classe` for the testing set observations.  
 
#### II.2. Expected out of sample error  
```{r cache=TRUE}
modFit.rf
```
As we saw above the Random forest algorithm delivers a clearly better accuracy. The estimated out of sample error (based on the out-of-bag (oob) error) is **0.49%**.  


### III. Predictions for the testing set provided   
Using the Random forest model build on the previous section, we predicted the answers to the observations provided on the testing set:  
```{r}
as.vector(predict(modFit.rf, testingRaw))
```


#### References  
[1] Ugulino, W.; Cardador, D.; Vega, K.; Velloso, E.; Milidiu, R.; Fuks, H. Wearable Computing: Accelerometers' Data Classification of Body Postures and Movements. Proceedings of 21st Brazilian Symposium on Artificial Intelligence. Advances in Artificial Intelligence - SBIA 2012. In: Lecture Notes in Computer Science. , pp. 52-61. Curitiba, PR: Springer Berlin / Heidelberg, 2012. ISBN 978-3-642-34458-9. DOI: 10.1007/978-3-642-34459-6_6. http://groupware.les.inf.puc-rio.br/har  
[2] Random Forests - http://www.stat.berkeley.edu/~breiman/RandomForests/cc_home.htm  
[2] Models Available in train By Tag - http://topepo.github.io/caret/bytag.html  
[3] R: how to total the number of NA in each col of data.frame - http://stackoverflow.com/questions/26273663/r-how-to-total-the-number-of-na-in-each-col-of-data-frame  
[4] Removing columns with missing values - http://stackoverflow.com/questions/7330915/removing-columns-with-missing-values   
[5] Remove an entire column from a data.frame in R - http://stackoverflow.com/questions/6286313/remove-an-entire-column-from-a-data-frame-in-r  
[6] Function for converting dataframe column type - http://stackoverflow.com/questions/11261399/function-for-converting-dataframe-column-type  
[7] How To Estimate Model Accuracy in R Using The Caret Package - http://machinelearningmastery.com/how-to-estimate-model-accuracy-in-r-using-the-caret-package/  
